Incremental tactics

### 2012-05-12 doc/papers/LFMTP2012/paper.tex
paper

When writing critical software, high assurance against bugs is an
absolute necessity. To certify the absence of bugs in a large body of
code given its specification, human reviewing is one option, but
certainly a failible one. Formal, mathematical proof that our program
respects a certain \emph{dependability claim} gives us the assurance
that ``its trustworthyness [\ldots] would presumably last for eternity
\cite{shao2010certified}''.

Certifying

\cite{miller2011proposal}

@@

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} {\lam x \lam y \s{\var x}}}
  \\
  &\quad\app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function $\lam x\lam y\s{\var
  x}$, and he could reuse the pieces of derivation just built. Let us
call them $\md_2$, $\md_3$ and $\md_4$. Provided there is an operator
$\get\md$ from the derivation to the program it types, what needs really
to be sent to the system is:
\begin{align*}
  &\infer{(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} {\get{\md_4}}}
    \app{\var{add}}\app{\get{\md_2}}{\get{\md_2}}
  )}
\ \text{,}
\end{align*}
that is, only the subterms of the program that changed. Derivations
$\md_2$ and $\md_4$ used to live in the empty context, we are then
free to use it in the new context where
$\var{add}:\cst{nat}\to\cst{nat}$, because System \sysname T enjoys
the \emph{weakening} property. Taking advantage of $\md_2$ and $\md_4$
the system succeeds with a new derivation made out of these, without
having to regenerate them because it knows that $\infer{\get\md} =
\md$. Let us call $\md_5$ the subderivation generated for $\tlam x
{\cst{nat}} \tlam y {\cst{nat}} \recb {\var x} {\var y}
{\get{\md_4}}$, $\md_6$ for the arguments
$\app{\get{\md_2}}{\get{\md_2}}$ and $\md_7$ for the call
$\app{\var{add}} {\get{\md_6}}$. Even if it occurs originally in the
scope of $\var{add}$, $\md_6$ doesn't depend on the hypothesis $\var
H$ that $\vdash \var{add}:\cst{nat}\to\cst{nat}$ because of the
\emph{strengthening} property. Contrarily, $\md_7$ does: each time we
want to reuse it, we will have to provide a proof of this fact.

The student then provides a definition for multiplication, and writes
the operation $(2 + 2)\times3$:
\begin{align*}
  \infer{}(
  &\letb{add}{\get{\md_5}}
  \\
  &\quad\letb{mul} { \tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
      \app{\var{add}}\app{\var x} {\var z}} }
  \\
  &\quad\quad\app{\var{mul}}\app
  {(\get{\gsubst{\md_7}{\msubst H {\infer{\var{add}}}}})}
  {(\s{\get{\md_2}})}
  )
  \ \text{.}
\end{align*}
Here again, he took care of reusing all derivations for equal
subterms. In particular, $\md_7$ expects a proof that the variable
$\var{add}$ is of type $\cst{nat}\to\cst{nat}$; he provides
$\infer{\var{add}}$ meaning: at this point we will have generated (or
reused in this case) the derivation of $\var{add}$, so just use it.

@@

% scale to big programs
% memoization ds le cas HO
%   nécessité de fcts inverses?
% séparer generation des diffs et vérification

@@

>>>>

@@

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} {\lam x \lam y \s{\var x}}}
  \\
  &\quad\app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function $\lam x\lam y\s{\var
  x}$, and he could reuse the pieces of derivation just built. Let us
call these $\md_2$, $\md_3$ and $\md_4$. Provided there is an operator
$\get\md$ from the derivation to the program it types, what needs
really to be sent to the system is:
\begin{align*}
  &\infer{(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} {\get{\md_4}}}
    \app{\var{add}}\app{\get{\md_2}}{\get{\md_2}}
  )}
\ \text{,}
\end{align*}
that is, only the subterms of the program that changed. Derivations
$\md_2$ and $\md_4$ used to live in the empty context, we are then
free to use it in the new context where
$\var{add}:\cst{nat}\to\cst{nat}$, because System \sysname T enjoys
the \emph{weakening} property. Taking advantage of the fact that the
judgement $\md_2$ and $\md_4$ proves is statically known, the system
succeeds with a new derivation made out of these, without having to
regenerate them because it knows that $\infer{\get\md} = \md$. Let us
call $\md_5$ the subderivation generated for $\tlam x {\cst{nat}}
\tlam y {\cst{nat}} \recb {\var x} {\var y} {\get{\md_4}}$, $\md_6$
for the arguments $\app{\get{\md_2}}{\get{\md_2}}$ and $\md_7$ for the
call $\app{\var{add}} {\get{\md_6}}$. Even if it occurs originally in
the scope of $\var{add}$, $\md_6$ doesn't depend on the hypothesis
$\var H$ that $\vdash \var{add}:\cst{nat}\to\cst{nat}$ because of the
\emph{strengthening} property. Contrarily, $\md_7$ does: each time we
want to reuse it, we will have to provide a proof of this fact.

The student then provides a definition for multiplication, and writes
the operation $(2 + 2)\times3$:
\begin{align*}
  \infer{}(
  &\letb{add}{\get{\md_5}}
  \\
  &\quad\letb{mul} { \tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
      \app{\var{add}}\app{\var x} {\var z}} }
  \\
  &\quad\quad\app{\var{mul}}\app
  {(\get{\gsubst{\md_7}{\msubst H {\infer{\var{add}}}}})}
  {(\s{\get{\md_2}})}
  )
  \ \text{.}
\end{align*}
Here again, he took care of reusing all possible subderivations. In
particular, $\md_7$ expects a proof that the variable $\var{add}$ is
of type $\cst{nat}\to\cst{nat}$; he provides $\infer{\var{add}}$
meaning: at this point we will have generated (or reused in this case)
the derivation of $\var{add}$, so just use it.

@@

Although this small example may seem anecdotal, the interest of a
mechanism for reusing results of derivation generation becomes clear
on very large programs: then the cost of rechecking well-typing of a
small modification amounts to recheking the path in the term from the
root to the changed part, and the changed part itself.

A type-checker is usually viewed as an algorithm returning a binary
answer (ok/error), specified by a set of inference rules. Seeing
type-checking as issueing a completely explicit typing derivation
enables to think of the relationship between the input and the
output. Feeding part of the output $\md_i$ back to the input $P_i$ is
a well-known technique for handling changes in the input: combined
with a way to automatically recognize already-computed input, it is
\emph{memoization}, a general-purpose technique for incremental
computation that has been extensively studied and combined with other
techniques \cite{pugh1989incremental, acar2003selective}. However, it
is best understood in the case of purely first-order data, and to our
best knowledge remains unexplored for higher-order structures
containing binders, like programs and proofs. In this case,

To simplify the problem,
we separate the task in:
\begin{itemize}
\item from a modified program, generate the \emph{diff} against the
  previous version, that is, observing and linking to unchanged
  derivations.
\item given a \emph{diff}, construct a new derivation and verify it.
\end{itemize}
The latter is the subject of this paper.

@@

% lambda-lifting
